Informative Oceanographic Sampling using Excursion Probabilities for Multivariate Random Fields

### 2019-08-26 main_new.tex
updates

% == Trygve Test
\usepackage{color, colortbl}
\definecolor{Gray}{gray}{0.9}
\newcommand{\kcomment}[1]{{\color{blue}{\{KR: #1\}}}}
\newcommand{\tofcomment}[1]{{\color{green}{\{TF: #1\}}}}

% put your definitions there:
\startlocaldefs
\endlocaldefs

@@

\address{Idiap \printead{e3}}
\affiliation{Uncertainty Quantification and Optimal Design group , Idiap Research Institute, Martigny, Switzerland}

\address{Department of Engineering Cybernetics,\printead{e4}}
\affiliation{Department of Engineering Cybernetics, NTNU}

@@

\begin{abstract}
Improving and optimizing oceanographic sampling is a crucial task for modern marine management and science faced with limited resources and highly dynamic processes. The combination of statistics and autonomous robotics provides opportunities for studying aspects related to experimental designs for effective characterization of oceanographic phenomena. In this paper methods are developed for efficient spatial sampling applied to the mapping of dynamic coastal processes by providing informative descriptions of spatial characteristics of ocean phenomena. Specifically, we define a design criterion based on reducing uncertainty in the excursion set of multivariate random fields, and derive analytical expressions for the expected Bernoulli variance reduction in the excursion set of spatial temperature and salinity variables. We show how these can be applicable in static designs and sequential designs, building on myopic and look-ahead sampling strategies. We use simulations to study the properties and differences between state-of-the-art approaches. Results from field deployments with an autonomous underwater vehicle are also presented, for a case study mapping the boundary of a river plume. The results demonstrate the ability of the method to effectively inform and execute data-driven environmental sampling.

@@

Monitoring the worlds oceans has gained increased importance in the light of the changing climate and increasing anthropogenic impact. A central problem for understanding these factors is the lack of representative data with sufficient resolution. Most of this \emph{undersampling} can be contributed to the large spatio-temporal variation on which ocean processes transpire, prompting the need for effective means of sampling. By \emph{sampling}, we refer to the design of observational strategies in the spatial domain, where the use of enabling autonomous platforms can be combined with statistical methods to pursuit measurements with high scientific relevance.

Traditional data collection at sea has typically been based on static buoys, lagrangian floats or ship-centric methods, with significant logistical limitations that directly impact coverage and sampling resolution. Modern methods using satellite remote-sensing provide large-scale coverage, but have limited resolution, are limited to the sensing the surface of the ocean, and is impacted by cloud cover. However, the advent of robust mobile robotic platforms \cite{Bellingham07} are making significant contributions to environmental monitoring and sampling. In particular autonomous underwater vehicles (AUVs), have made significant strides in advancing the state of sampling and consequently have made robotics an integral part of ocean observation in large part driven by using statistical methods in design and data assimilation \citep{das11b,Graham2013,Das2015,das15,fossuminformation,fossum18b}. Others include \cite{wikle2013modern} who focus on hierarchical statistical models; \cite{sahu2008space} study spatio-temporal models for sea surface temperature and salinity data, while \cite{mellucci2018oceanic} look at the statistical prediction of features using an underwater glider. Common to these approaches is the use of information-based methods to infer environmental sampling and a few important characteristics should be mentioned, namely:

\begin{itemize}
    \item Decisions on where to sample happen sequentially, meaning recent information can be assimilated and used to improve statistical models and reduce uncertainty associated with the  choices. 
    \item As a priori information about the state of the ocean is very uncertain, there lies a substantial efficiency gain in reacting to information obtained from measurements taken \emph{in-situ}, this is usually referred to as the \emph{adaptivity gap}.
    \item For this type of sampling problem, the number of choices (i.e. locations, paths, and candidate designs) makes the problem combinatorially large, creating a trade-off between optimization (finding the optimal design) and computability (arriving at a solution in reasonable time), which has to be considered in practice.
\end{itemize}{}

The combination of statistical tools and robotic platforms is a natural symbiosis which enables information-based sensing. Central to this is the ability to model spatially correlated variables and provide formal measures of uncertainty. By far the most practical models use Gaussian processes (GPs), which will also be used in our formulation, in large  part because this approach allows efficient implementation and evaluation, enabling prioritization of sampling efforts to be formalized into decision making systems on-board a robotic platform.

@@

Central to understanding the changes taking place in the upper water-column is the understanding of biophysical interaction  driven by an agglomeration of physical forcing (wind, topography, bathymetry, tidal influences and local geography such as fresh water inflows) mixed with incipient micro-biology driven by planktonic and coastal anthropogenic input, such as pollution and agricultural runoff, which results in a range of ecosystem phenomena (e.g. blooms, anoxic zones, plumes) ---all influences with significant impact both directly and indirectly on society. Reverberating the requirement of more effective ocean sensing practices, renowned oceanographer Walter Munk referred to the 20\textsuperscript{th} century as the \emph{century of undersampling} \citep{munk2002}, pointing out that lack of sufficient sampling resolution in time and space, as a critical issue for characterizing ocean phenomena. Still today, decision makers struggle with limited observational efforts sometimes leading to severe environmental problems. The recent (May 2018) outbreak of harmful algal blooms in Northern Norway\footnote{Report from the Institute of Marine Research:https://bit.ly/2KZGJfI}, with losses exceeding 12 million tons of salmon valued at \$ 80 million (Norwegian Directorate of Fisheries), is a case in point in how adequate sampling efforts could have provided a more accurate warning assessment. These challenges of undersampling have been driving ocean sensing practices towards more networked and distributed use of robotic assets in order to fill the gaps in our understanding (See Fig. \ref{fig:envir}), where the surveying is guided by a model and the measurements made onboard the robot are used to update the model.

\begin{figure}[!h] 
  \centering 
  \subfigure[Modern ocean sensing network, where robotic assets augment ship based measurements to increase coverage of the phenomena.]{\includegraphics[width =
    0.48\textwidth]{Figures/envir.pdf}\label{fig:envir1}}
  \hfill
  \subfigure[Data-driven sampling under the sense-plan-act paradigm.]{\includegraphics[width =
    0.48\textwidth]{Figures/adaptive_cycle.pdf}\label{fig:sense-plan-act}}
  \caption{Ocean observation is moving away from single-ship sampling towards more autonomous and collaborative networked operations in order to resolve the numerous processes and their interaction.} \label{fig:envir}
\end{figure}

Focusing on coupling statistics and robotic sampling, our attention is in this work towards spatial characterization of a frontal systems, and more specifically fronts created by river plume systems. Fig. \ref{fig:nidelven} shows the survey area (the outlet of the Nidelva river, Trondheim, Norway) taken as cold freshwater enters from the river creating a strong separation gradient in both temperature and salinity. Because of the local topography and the Coreolis force \citep{coriolis1835memoire} the cold fresh water tends to flow near land to the east. Depending on the river discharge, tidal effects, wind and temperature differences, this boundary in water masses will get distorted. Prior knowledge about the location and evolution of these features is therefore highly uncertain, making deterministic planning sub-optimal.

\begin{figure}[!h] 
\centering 
\includegraphics[width=0.85\textwidth]{Figures/pictures/c-updated.pdf}
\caption{Frontal patterns - Nidelven, Trondheim, Norway.}
\label{fig:nidelven}
\end{figure}

@@

River plumes belong to a class of ocean processes that are local and usually act on smaller \emph{sub-mesoscale}, from 10 m -- 5 km where spatial lateral variability tends to dominate. Remote sensing and synthetic ocean models find it challenging to provide detailed at this scale \citep{Lermusiaux:2006} and the principal way to resolve for fine resolution is by direct observation. At larger \emph{mesoscale} ($>50$ km), both three-dimensional space and time dynamics are important and can shift substantially, while in the case of river plumes one can often limit scope to cover only lateral spatial elements ([north, east] domain). Moreover, for this case, time-effects can be regarded as static when acquiring AUV data over a short time window and limited region. However, the methodological framework can be extended to higher dimensional situations in a similar manner.

>>>>

% == Trygve Test
\usepackage{color, colortbl}
\definecolor{Gray}{gray}{0.9}
\newcommand{\kcomment}[1]{{\color{red}{\{KR: #1\}}}}
\newcommand{\tofcomment}[1]{{\color{green}{\{TF: #1\}}}}
\newcommand{\kc}{\kcomment}
% put your definitions there:
\startlocaldefs
\endlocaldefs

@@

\address{Idiap \printead{e3}}
\affiliation{Uncertainty Quantification and Optimal Design group, Idiap Research Institute, Martigny, Switzerland}

\address{Department of Engineering Cybernetics,\printead{e4}}
\affiliation{Department of Engineering Cybernetics, NTNU \&\\ Underwater Systems and Technology Laboratory, Faculty of Engineering, University of Porto, Portugal}

@@

\begin{abstract}
  Improving and optimizing oceanographic sampling is a crucial task for
  modern marine management and science faced with limited resources to
  understand highly dynamic processes in the water-column. The
  combination of statistics and autonomous
  robotics provides opportunities for studying aspects related to
  experimental designs for effective characterization of oceanographic
  phenomena. In this work we develop methods for efficient spatial
  sampling applied to the mapping of dynamic coastal processes by
  providing informative descriptions of spatial characteristics of ocean
  phenomena. Specifically, we define a design criterion based on
  reducing uncertainty in the excursion set of multivariate random
  fields, and derive analytical expressions for the expected Bernoulli
  variance reduction in the excursion set of spatial temperature and
  salinity variables. We show how these can be applicable in static
  designs and sequential designs, building on myopic and look-ahead
  sampling strategies. We use simulations to study the properties and
  differences between state-of-the-art approaches. Results from field
  deployments with an autonomous underwater vehicle are also presented,
  for a case study mapping the boundary of a river plume. The results
  demonstrate the ability of the method to effectively inform and
  execute data-driven environmental sampling.

@@

Monitoring the worlds oceans has gained increased importance in the
light of the changing climate and increasing anthropogenic impact. A
central problem for understanding these factors is the lack of
representative data with sufficient resolution. Most of this
\emph{undersampling} can be attributed to the large spatio-temporal
variations on which ocean processes transpire, prompting the need for
effective means of sampling. By \emph{sampling}, we refer to the design
of observational strategies in the spatial domain, where the use of
autonomous platforms can be combined with statistical methods to pursue
measurements with high scientific relevance.

Traditional data collection at sea has typically been based on static
buoys, lagrangian floats or ship-based methods, with significant
logistical limitations that directly impact coverage and sampling
resolution. Modern methods using satellite remote-sensing provide
large-scale coverage, but have limited resolution, are limited to
sensing the surface of the ocean, and is impacted by cloud cover. The
advent of robust mobile robotic platforms \cite{Bellingham07} has
resulted in significant contributions to environmental monitoring and
sampling. In particular autonomous underwater vehicles (AUVs), have made
strides in advancing the state of sampling and consequently have made
robotics an integral part of ocean observation in large part by our own
work % driven by
% using statistical methods in design and data assimilation
\citep{das11b,Graham2013,Das2015,das15,fossuminformation,fossum18b}.
Others include \cite{wikle2013modern} focusing on hierarchical
statistical models, \cite{sahu2008space} which studies spatio-temporal
models for sea surface temperature and salinity data, while
\cite{mellucci2018oceanic} looks at the statistical prediction of
features using an underwater glider. Common to these approaches is the
use of information-based methods to infer environmental sampling. A few
important characteristics should be mentioned, namely:

\begin{itemize}
\item Decisions on where to sample happen sequentially, i.e.  how
  recently acquired information can be assimilated and used to improve
  statistical models and reduce uncertainty associated with where and
  when to make measurements.
\item As a priori information about the state of the ocean is very
  uncertain, there lies a substantial efficiency gain in reacting to
  information obtained from measurements taken \emph{in-situ}, this is
  usually referred to as the \emph{adaptivity gap} \kc{citation
    needed}.
\item For such types of sampling problems, the number of choices (i.e.
  locations, trajectories, and candidate designs) makes the problem
  combinatorially large, creating a trade-off between optimization
  (finding the optimal design) and computability (arriving at a
  solution in reasonable time), which has to be considered in
  practice.
\end{itemize}{}

The combination of statistical tools and robotic platforms is a
natural symbiosis which enables information-based sensing. Central to
this is the ability to model spatially correlated variables and
provide formal measures of uncertainty. By far the most practical
models use Gaussian processes (GPs), which will also be used in our
formulation, in large part because it allows efficient implementation
and evaluation, in turn enabling prioritization of sampling efforts to
be formalized into decision making systems on-board a robotic
platform.

@@

Central to understanding the changes taking place in the upper
water-column is the understanding of biophysical interaction driven by
an agglomeration of physical forcing (wind, topography, bathymetry,
tidal influences and local geography such as fresh water inflows)
mixed with incipient micro-biology driven by planktonic and coastal
anthropogenic input, such as pollution and agricultural runoff. These
often result in a range of ecosystem phenomena (e.g. blooms, anoxic
zones, plumes) all with significant impact both directly and
indirectly on society. Reverberating the requirement of more effective
ocean sensing practices, Walter Munk referred to the
20\textsuperscript{th} century as the \emph{century of undersampling}
\citep{munk2002}, pointing out that lack of sufficient sampling
resolution in time and space, as a critical issue for characterizing
ocean phenomena. Even today policy makers struggle with limited
observational efforts sometimes leading to severe environmental
problems such as the recent (May 2018) outbreak of harmful algal
blooms in Northern Norway\footnote{Report from the Institute of Marine
  Research:https://bit.ly/2KZGJfI}, with losses exceeding 12 million
tons of salmon valued at \$ 80 million (Norwegian Directorate of
Fisheries). This is a case in point in how adequate sampling efforts
could have provided a more accurate assessment and prior
warning. These challenges related to undersampling have been driving
ocean sensing practices towards more autonomous, networked and
distributed use of robotic assets in order to fill the gaps in our
understanding (See Fig. \ref{fig:envir}), where the surveying is
guided by a dynamical ocean model and the measurements made onboard
the robot are often used to update the model.

\begin{figure}[!h] 
  \centering 
  \subfigure[Illustration of an ocean sensing network, where robotic
  assets augment ship based measurements to increase coverage and
  resolution of phenomena of scientific interest.]{\includegraphics[width =
    0.48\textwidth]{Figures/envir.pdf}\label{fig:envir1}}
  \hfill
  \subfigure[Data-driven sampling under the sense-plan-act paradigm.]{\includegraphics[width =
    0.48\textwidth]{Figures/adaptive_cycle.pdf}\label{fig:sense-plan-act}}
  \caption{Ocean observation is moving away from single-ship sampling
    towards more autonomous and collaborative networked operations in
    order to resolve the numerous bio-geochemical processes and their
    interactions.}
\label{fig:envir}
\end{figure}

Our attention in this work is towards spatial characterization of a
frontal system, and more specifically fronts generated by river
plumes. Fig. \ref{fig:nidelven} shows the survey area (the outlet of
the Nidelva river, Trondheim, Norway) taken as cold freshwater enters
from the river creating a strong gradient in both temperature and
salinity. Because of the local topography and the Coreolis force
\citep{coriolis1835memoire} the cold fresh water tends to flow near
land to the east. Depending on the river discharge, tidal effects,
wind and temperature differences, this boundary in water masses often
gets distorted. Prior knowledge about the location and evolution of
these features is therefore highly uncertain, making deterministic
planning challenging.

\begin{figure}[!h] 
\centering 
\includegraphics[width=0.85\textwidth]{Figures/pictures/c-updated.pdf}
\caption{Frontal patterns off of the Nidelva river, Trondheim, Norway.}
\label{fig:nidelven}
\end{figure}

@@

River plumes belong to a class of ocean processes that are local and
at smaller \emph{sub-mesoscale}, from 10 m -- 5 km where spatial
lateral variability tends to dominate. Remote sensing and synthetic
ocean models find it challenging to provide detail at this scale
\citep{Lermusiaux:2006} and the principal way to resolve for fine
resolution is by direct observation. At larger \emph{mesoscale}
($>50 km^2$), both three-dimensional space and time dynamics are
important and can shift substantially, while in the case of river
plumes one can often limit scope to cover only lateral spatial
elements ([north, east] domain). Moreover, for this case, time-effects
can be regarded as static when acquiring AUV data over a short time
window and limited region. However, the methodological framework can
be extended to higher dimensional situations in a similar manner.
